---
title: "Analysis of U.S. National Disease Death Trends (2014-2021)"
author: "Patrick Cook"
date: "8/13/2021"
output: 
  html_document:
    # df_print: paged
    keep_md: True
---

```{r, global.options, include = FALSE}
knitr::opts_chunk$set(
#    fig.width   = 12,          # the width for plots created by code chunk
#    fig.height  = 12,          # the height for plots created by code chunk
    dpi         = 200,         # set dpi for hd screen
    fig.align   = 'center',    # align graphics in the final doc
    out.width   = "90%",       # Set output (html) to % of window
    out.height  = "90%",       # Set output (html) to % of window
    # fig.path    = 'images/', # file path to graphics storage directory
    echo        = TRUE,       # do not display code in the code chunk
    message     = TRUE,       # do not display any messages generated by code
    strip.white = TRUE,        # remove white spaces at beg or end of code chunk
    warning     = TRUE,       # do not display any warning messages
    results     = TRUE)       # show text output
```


```{r , echo=FALSE}

# Update/Create cleaned data set and import variables and functions
  source("clean_and_merge.R")

```

**Note**: Initial data sets were cleaned and merged prior to this analysis. The 
original data sets' structure and cleaning steps may be found in the file 
[us_state_death_trends_wrangling](us_state_death_trends_wrangling.Rmd).

### Introduction and Objectives

This report explores the Center for Disease Control's (CDC) Weekly Morbidity and 
Mortality data from 2014 through the **------>1st quarter of 2021<--------**. 
This data analysis is focused on National level weekly death counts for specific 
causes at the United States. A 2nd report will analyze specific state and local 
regions. The data is voluntary and not guaranteed to be reported in a timely or 
regular basis. Therefore, the most recent data (quarter) will be may be 
incomplete and unreliable for analysis and insights. 

More information on the original data sets can be found at CDC's website using 
the following links: 

- [2014-2015 MMWR Data Set](https://data.cdc.gov/NCHS/Weekly-Counts-of-Deaths-by-State-and-Select-Causes/3yf8-kanr){target="_blank"}
- [2020-2021 MMWR Data Set](https://data.cdc.gov/NCHS/Weekly-Provisional-Counts-of-Deaths-by-State-and-S/muzy-jte6){target="_blank"}

**Note**: The most current data can be downloaded using the links above. To use the 
data sets without any code modifications, the user will need to:

1. Place the data two sets in a subdirectory named "data", and
2. Rename the data sets as "weekly_2014_2019" and "weekly_2020_2021."

### Cleaned data set structure  

```{r}

str(mmwr_1421_df)

```


