---
title: "DRAFT Analysis of U.S. National Disease Death Trends (2014-2021)"
author: "Patrick Cook"
date: "8/13/2021"
output: 
  html_document:
    df_print: paged
    # keep_md: True
---

```{r, global.options, include = FALSE}
knitr::opts_chunk$set(
#    fig.width   = 12,        # the width for plots created by code chunk
#    fig.height  = 12,        # the height for plots created by code chunk
    dpi         = 200,        # set dpi for hd screen
    fig.align   = 'center',   # align graphics in the final doc
    out.width   = "90%",      # Set output (html) to % of window
    out.height  = "90%",      # Set output (html) to % of window
    fig.path    = 'images/',  # file path to graphics storage directory
    echo        = TRUE,       # do not display code in the code chunk
    message     = TRUE,       # do not display any messages generated by code
    strip.white = TRUE,       # remove white spaces at beg or end of code chunk
    warning     = TRUE,       # do not display any warning messages
    results     = TRUE)       # show text output
```


```{r , echo=FALSE}

# Update/Create cleaned data set and import variables and functions
  source("clean_and_merge.R")

```

# Draft

**Note**: Initial data sets were cleaned and merged prior to this analysis. The 
original data sets' structure and cleaning steps may be found in the file 
[us_state_death_trends_wrangling](us_state_death_trends_wrangling.Rmd).

## Introduction and Objectives  

This report explores the Center for Disease Control's (CDC) Weekly Morbidity and 
Mortality data from 2014 through the **------>1st quarter of 2021<--------**. 
**This data analysis is focused on National level weekly death counts for specific 
causes at the United States**. A 2nd report will analyze specific state and local 
regions. The data is voluntary and not guaranteed to be reported in a timely or 
regular basis. Therefore, the most recent data (quarter) will be may be 
incomplete and unreliable for analysis and insights. 

More information on the original data sets can be found at CDC's website using 
the following links: 

- [2014-2015 MMWR Data Set](https://data.cdc.gov/NCHS/Weekly-Counts-of-Deaths-by-State-and-Select-Causes/3yf8-kanr){target="_blank"}
- [2020-2021 MMWR Data Set](https://data.cdc.gov/NCHS/Weekly-Provisional-Counts-of-Deaths-by-State-and-S/muzy-jte6){target="_blank"}

**Note**: The most current data can be downloaded using the links above. To use the 
data sets without any code modifications, the user will need to:

1. Place the data two sets in a subdirectory named "data", and
2. Rename the data sets as "weekly_2014_2019" and "weekly_2020_2021."


```{r, echo=FALSE, results='hide'}
# Create United States subset
us_deaths_df <- mmwr_1421_df[Location == "United States", ]

# Dropping unused levels (only United States occurs). 
us_deaths_df$Location <- droplevels(us_deaths_df$Location)

any(is.na(us_deaths_df))
any(is.infinite(us_deaths_df[ , names(us_deaths_df)]))

```


## Data Structure  

### Observations and features  

#### Verify United States data set is smooth at merge (2019 to 2020).
#### **Note**: State level data will be verified in individual State analysis.   

```{r}

melt(us_deaths_df[ Week_End_Date > as.Date("2019-09-01") & 
                   Week_End_Date < as.Date("2020-03-31"),
                   Week_End_Date:Abnormal_Finding],
     id.vars = "Week_End_Date") %>%
  
  ggplot(aes(x=Week_End_Date, y = value, group=variable)) +
    geom_point() +
    geom_vline(xintercept=as.Date("2020-01-01"), linetype="dotted") +
    facet_wrap(~ variable, scales = 'free_y') + 
    theme_bw()

```

There are vertical gaps (jumps) seen at the intersection of the two data sets 
for: 
- Influenza and Pneumonia
- Other Respiratory
- Abnormal Finding

After the jumps, the data settles into a pattern consistent with the trend of 
the previous years' data. These gaps are likely related to Covid-19 cases that 
were undiagnosed due to Covid-19 testing not being available until March of 2020 
and not being widely available (without restricted use) until May of 2020. In 
addition, there was not a Covid-19 diagnosis of death code available in the 
United States on January 1st of 2020 and therefore any deaths would 
have been diagnosed as another general respiratory category such as these three.

In addition to the gaps, there are notable peaks at the merge point (date) of 
the two data sets. These peaks are generally smooth before and after, indicating 
a local anomaly with a true long-term pattern. 

The analysis will comprise mostly of averages and comparisons of year to year 
descriptive statistics. Therefore, 1-2 week gaps and trends will not effect the 
analysis. 

#### First three observations (chronologically) of the data set.  


```{r }
head(us_deaths_df, 3)

```

#### Last three observations (chronologically) of the data set.  

```{r }

tail(us_deaths_df, 3)

```

```{r }

str(us_deaths_df)

```



After preprocessing and cleaning, the United States (U.S.) subset of data used 
in this analysis accounts for 398 observations and 18 features which includes 
categorical location data, chronological date and week of year information, and 
integer weekly disease death data. The earliest week is the 1st week of January 
2014. The most current as of this analysis is the week ending January 24, 2021. 
Next, is the  data set's summary statistics in the U.S. subset.


#### Summary statistics  

```{r, message=FALSE, warning=FALSE }

describe(us_deaths_df[, Location:Week_End_Date])


```

There are 398 observations beginning with the week ending date of 
01-04-2021 to 08-14-2021. The years 2014 and 2020 have a week number 53 due to 
leap year's extra day and the last week of the year being splitting the 
following year, This 53rd week will not significantly affect the results of the 
analysis due to the analysis using year long and quarterly averaging. Year 2021 
only has 32 weeks. Any analysis after week 32 will show 2020 to previous year 
comparisons. After further review of the data, analysis will be limited to the 
end of the 2nd quarter of 2021 (see "Examining possible reporting delays in last 
quarter of data below"). 

#### Cause of weekly deaths summary statistics
#### **Note** the Covid-19 data contains fill values of zero for years 2014 through 2019 (inclusive). 

```{r, message=FALSE, warning=FALSE }

as.data.frame(describeBy(us_deaths_df[ , Natural:Abnormal_Finding]))


```

#### Summary statistics for Covid features without fill values (2020-2021)  

```{r, message=FALSE, warning=FALSE }

as.data.frame(describeBy(us_deaths_df[ Year > 2019, 
                                       .(Covid_19_Multi, Covid_19)]))

```




#### Distributions 2014-2021 (Covid deaths are zero prior to 2020)  

```{r, fig.width=10, fig.height=8}
us_deaths_df[ , Natural:Abnormal_Finding]%>%
  hist()
```

Most of the distributions above show data has a normal-like distribution with 
many having right-skew. The Covid categories are effected by the zeros from 
years before 2020 when there was not a Covid-19 diagnosis. These will be removed 
by limiting the data to the years 2020 and greater. The cancer and abnormal 
categories will benefit from a larger bin size. 


#### Distributions 2014-2021 (Covid distibutions are years 2020-2021)  

```{r, fig.width=8, fig.height=6}
par(mfrow=c(2,2))

  hist(us_deaths_df[ , Cancer], breaks = 22,
       xlim = c(5000, 14000),
       main = "Cancer", 
       xlab = "Weekly Cancer Deaths") 
  hist(us_deaths_df[ Year > 2019, Covid_19_Multi], breaks = 13,
              xlim = c(0, 30000),
       main = "Covid-19 Comorbidity", 
       xlab = "Weekly Covid-19 Comorbidity Deaths")
  hist(us_deaths_df[ Year > 2019, Covid_19], breaks = 12,
       xlim = c(0, 25000),
       main = "Covid-19 Singular Cause", 
       xlab = "Weekly Covid-19 Deaths")
  hist(us_deaths_df[ , Abnormal_Finding], breaks = 50,
       xlim = c(100, 3000),
       main = "Abnormal Finding", 
       xlab = "Weekly Abnormal Finding Deaths")

par(mfrow=c(1,1))

```

The cancer and abnormal findings categories show normal-like data distributions 
with cancer having possible outliers to the left and abnormal findings having 
significant outliers to the right. Outliers are expected with abnormal findings 
due to the diagnosis being itself an outlier from all other diagnoses. Cancer 
would be expected to be consistent with no weeks being significantly different 
than others. Therefore, cancer will require further analysis.

#### Exploring outliers in cancer data  


```{r, fig.width=4, fig.height=4}

plot(us_deaths_df$Week_End_Date, us_deaths_df$Cancer)

```

The outliers occur during the last 3 weeks of the data and possibly extend 
further. This is likely due to reporting delays and will affect all categories.  


#### Examining possible reporting delays in last quarter of data  

```{r, fig.width=14, fig.height=14}

melt(us_deaths_df[ ,
                   Week_End_Date:Abnormal_Finding],
     id.vars = "Week_End_Date") %>%
  
  ggplot(aes(x=Week_End_Date, y = value, group=variable)) +
    geom_point() +
    geom_vline(xintercept=as.Date("2021-07-01"), linetype="solid", color = "blue") +
    facet_wrap(~ variable, scales = 'free_y') + 
    theme_bw()

```

All categories except abnormal findings and Covid-19 related categories show the 
same pattern of outliers likely due to delayed reporting. Some categories, such 
as Alzheimers, show possible delayed reporting since the end of the 2nd quarter 
(blue line at 2021-07-01). To prevent possible errors, the analysis will be 
limited to data through the 2nd quarter of 2021 (2021-06-30). 


#### Distributions after dropping 2021 3rd quarter data  


```{r, fig.width=10, fig.height=8}
# Limit data set through 2nd quarter 2021 due to delayed reporting.
us_deaths_df <- us_deaths_df[Week_End_Date < as.Date("2021-07-01")]

us_deaths_df[ , Natural:Abnormal_Finding]%>%
  hist()
```




```{r, fig.width=8, fig.height=6}
par(mfrow=c(2,2))

  hist(us_deaths_df[ , Cancer], breaks = 22,
       xlim = c(10000, 13000),
       main = "Cancer", 
       xlab = "Weekly Cancer Deaths") 
  hist(us_deaths_df[ Year > 2019, Covid_19_Multi], breaks = 13,
              xlim = c(0, 30000),
       main = "Covid-19 Comorbidity", 
       xlab = "Weekly Covid-19 Comorbidity Deaths")
  hist(us_deaths_df[ Year > 2019, Covid_19], breaks = 12,
       xlim = c(0, 25000),
       main = "Covid-19 Singular Cause", 
       xlab = "Weekly Covid-19 Deaths")
  hist(us_deaths_df[ , Abnormal_Finding], breaks = 50,
       xlim = c(100, 3000),
       main = "Abnormal Finding", 
       xlab = "Weekly Abnormal Finding Deaths")

par(mfrow=c(1,1))

```

#### Chronological data after dropping 2021 3rd quarter data  

```{r, fig.width=14, fig.height=14}

melt(us_deaths_df[ ,
                   Week_End_Date:Abnormal_Finding],
     id.vars = "Week_End_Date") %>%
  
  ggplot(aes(x=Week_End_Date, y = value, group=variable)) +
    geom_point() +
    facet_wrap(~ variable, scales = 'free_y') + 
    theme_bw()

```

### Seasonal Patterns Seen in Non-seasonal Causes

Seasonality trends occur in all cause of death categories. This is expected in 
respiratory causes of death. Non-respiratory diseases, such as cancer, 
Alzheimers and diabetes, unexpectedly show seasonality patterns with higher 
weekly death counts in fall and winter. The next figure examines the 
chronological correlations of the causes of deaths for the 2020 and 2021 data.
Earlier years show similar correlations. 


#### Seasonal chronological correlations between causes of weekly deaths

```{r, fig.width=14, fig.height=14}
# psych::pairs.panels(us_deaths_df[ , Natural:Abnormal_Finding], scale = TRUE)

```

```{r, fig.width=14, fig.height=12}
psych::pairs.panels(us_deaths_df[ Year > 2019 , Natural:Abnormal_Finding], scale = TRUE)

```

 Strong correlations possibly exist between natural deaths and heart, cerebral 
 (brain), Alzheimers, diabetes and Covid-19 causes of weekly deaths. The heart 
 weekly deaths category show a moderate to strong correlation with all 
 categories except abnormal findings. Cerebral (brain) weekly deaths have a 
 similar relationship excluding lower respiratory and abnormal findings. As 
 expected, the respiratory causes of death categories show correlations.
 
 #### Causes of death categories with strongest possible correlations  
 

```{r, fig.width=10, fig.height=10}
psych::pairs.panels(us_deaths_df[ Year > 2019,
                                  .(Natural, Heart, Brain, Alzheimer, Diabetes, 
                                    Covid_19_Multi, Covid_19)], 
                    scale = TRUE)

```

The seasonal patterns could be the result of patients having heart disease and 
becoming infected with a seasonal virus causing additional stress on the immune 
system. In addition, natural death is an age based cause of death category. If 
an elderly patient dies of unknown or comorbid causes, the cause of death will 
be coded as natural cause of death. 

Though not in the scope of this analysis, these seasonal pattern is interesting 
in that it points to a possible virus as a cause of death and would benefit from 
further study. Additional data, such as age and any secondary cause of death 
data would be beneficial and is available at the National Center for Health 
Statistics. 

#### Long-term trends (2014-2021)

```{r}
names(us_deaths_df)
```


```{r, fig.width=14, fig.height=12}

melt(us_deaths_df[ ,
                   -c("Location", "Year", "Week", "Covid_19", "Covid_19_Multi")],
     id.vars = "Week_End_Date") %>%
  
  ggplot(aes(x=Week_End_Date, y = value, group=variable)) +
    geom_point() +
    geom_smooth(se = FALSE,
                size = 1.5,
                col=mycolors[24]) +
    facet_wrap(~ variable, scales = 'free_y') + 
    theme_bw()

```

```{r, fig.width=7, fig.height=5}

melt(us_deaths_df[ Year > 2019,
                   c("Week_End_Date", "Covid_19_Multi", "Covid_19")],
     id.vars = "Week_End_Date") %>%
  
  ggplot(aes(x=Week_End_Date, y = value, group=variable)) +
    geom_point() +
    geom_smooth(method = "lm",
                se = FALSE,
                size = 1.5,
                col=mycolors[24]) +
    facet_wrap(~ variable, scales = 'free_y') + 
    theme_bw()

```

# End of file  



