---
title: "U.S. States' Disease Death Trends (2014-2021)"
author: "Patrick Cook"
date: "7/21/2021"
output: 
  html_document:
    df_print: paged
---

```{r, global.options, include = FALSE}
knitr::opts_chunk$set(
#    fig.width   = 12,          # the width for plots created by code chunk
#    fig.height  = 12,          # the height for plots created by code chunk
    dpi         = 200,         # set dpi for hd screen
    fig.align   = 'center',    # align graphics in the final doc
    out.width   = "90%",       # Set output (html) to % of window
    out.height  = "90%",       # Set output (html) to % of window
    # fig.path    = 'images/', # file path to graphics storage directory
    echo        = FALSE,       # do not display code in the code chunk
    message     = FALSE,       # do not display any messages generated by code
    strip.white = TRUE,        # remove white spaces at beg or end of code chunk
    warning     = FALSE,       # do not display any warning messages
    results     = TRUE)       # show text output
```


```{r , include=FALSE}
# Install packages
pacman::p_load(pacman, data.table, dplyr, tidyr, knitr, ggplot2, psych, cowplot, 
               corrplot, stringr, Hmisc, RColorBrewer)

# -----------------------------------------------------------------------------
# -------------- Set color palettes and sorting/helper variables --------------
# -----------------------------------------------------------------------------

# Diverging Palette: Modified from Brewer"Dark2", "Set1" and "Paired" palettes.
mycolors <- c("#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#00aaff", "#E6AB02", 
              "#A6761D", "#666666", "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", 
              "#FF7F00", "#FFFF33", "#A65628", "#F781BF", "#999999", "#66A61E", 
              "#B2DF8A", "#33A02C", "#FB9A99", "#E31A1C", "#FDBF6F", "#FF7F00", 
              "#CAB2D6", "#6A3D9A", "#FFFF99", "#B15928")

# create list of system greys (every other shade)
mygreycolors <- sprintf('grey%s', seq(0, 100, 2))

# Order string years [add current years as included]
ordered_years <- c("2014", "2015", "2016",
                   "2017", "2018", "2019", "2020", "2021")

# Some years are given 53 weeks in the year (usually leap years)
ordered_weeks <- seq(1:53)

# Order of states and special regions by population, greatest to least.
# Used for viewing states with similar population sizes and graphic ordering
state_by_pop <- c("California", "Texas", "Florida", "New York", "Pennsylvania", 
                  "Illinois", "Ohio", "Georgia", "North Carolina", "Michigan", 
                  "New Jersey", "Virginia", "New York City", "Washington", 
                  "Arizona", "Massachusetts", "Tennessee", "Indiana", 
                  "Missouri", "Maryland", "Wisconsin", "Colorado", "Minnesota", 
                  "South Carolina", "Alabama", "Louisiana", "Kentucky", 
                  "Oregon", "Oklahoma", "Connecticut", "Utah", "Puerto Rico", 
                  "Iowa", "Nevada", "Arkansas", "Mississippi", "Kansas", 
                  "New Mexico", "Nebraska", "West Virginia", "Idaho", "Hawaii", 
                  "New Hampshire", "Maine", "Montana", "Rhode Island", 
                  "Delaware", "South Dakota", "North Dakota", "Alaska", 
                  "District of Columbia", "Vermont", "Wyoming"
                  )

# End helper Variables

```

```{r , include=FALSE}
# -----------------------------------------------------------------------------
# --------------------------------- Functions ---------------------------------
# -----------------------------------------------------------------------------



# End Functions

```

### Introduction and Objectives

This report explores the Center for Disease Control's (CDC) Weekly Morbidity and 
Mortality data from 2014 through the 1st quarter of 2021. 
The data includes weekly death counts for specific causes in the United States. 
The data also includes weekly death counts for individual states and specific 
areas such as New York City, Puerto Rico and Washington, D.C. The data is 
voluntary and not guaranteed to be reported in a timely or regular basis. 
Therefore, the most recent data (quarter) will be incomplete and unreliable for 
analysis and insights. More information on the data set can be found at CDC's 
website using the following links: 

- [2014-2015 MMWR Data Set](https://data.cdc.gov/NCHS/Weekly-Counts-of-Deaths-by-State-and-Select-Causes/3yf8-kanr){target="_blank"}
- [2020-2021 MMWR Data Set](https://data.cdc.gov/NCHS/Weekly-Provisional-Counts-of-Deaths-by-State-and-S/muzy-jte6){target="_blank"}

Note: The most current data can be downloaded using the links above. To use the 
data sets without any code modifications, the user will need to:

1. Place the data two sets in a subdirectory named "data", and
2. Rename the data sets as "weekly_2014_2019" and "weekly_2020_2021."


```{r , include=FALSE}

# import data sets (included in data folder)

mmwr_1419_df = read.csv("data/weekly_2014_2019.csv",
                   header = TRUE,
                   stringsAsFactors = FALSE)

mmwr_2021_df = read.csv("data/weekly_2020_2021.csv",
                   header = TRUE,
                   stringsAsFactors = FALSE)

```


#### Objective

The objective of the project is to analyze national and state level weekly cause
of death data available from the Centers for Disease Control (CDC) to identify 
specific diseases with the highest **growth** in deaths so that the information
may guide long-term and short-term planning. This planning will allow healthcare
systems to redirect resources, reducing financial, material, facility and labor 
cost wastage as well as identifying and addressing the cause for the increased 
deaths before the disease becomes a leading contributor of deaths.

### Initial Data Structure Exploration

```{r }
str(mmwr_1419_df)

```

```{r }
str(mmwr_2021_df)

```

```{r }
head(mmwr_1419_df, 3)
head(mmwr_2021_df, 3)
```

#### Observations above

1.  The 2014-2019 data set has 16,902 records with 30 features. The 2020-2021 
data contains 4,428 records with 35 features. The difference in features is due 
to the addition of two Covid-19 categories in the 2020-2021 data set.

2. Both data sets have flag columns that identify weekly death counts that are 
from 1 to 9 deaths that have been redacted due to privacy issues. The records 
with missing values are currently all due to the privacy flag. These flags 
columns are not needed for the analysis of growth rates and will be dropped. The
missing values will not be replaced due to the influence a zero or interpolated 
value would have on statistical calculations and patterns seen in plots. The 
missing values do still have value in that they indicate a positive weekly death 
count below 10. 

2. Both data sets contain coded feature names based on International Classification 
of Diseases (ICD-10) that will be renamed to shorter names based on affected 
organ or two word disease name.

3. The dates in both data sets are listed as characters and are not in the 
global standard format. Data type will be changed and put into ISO 8601 format.

4. The jurisdictions will be renamed to "Location" and changed to a factor.

5. Other features such as year and week may be changed to ordered factors as 
needed.

Both data sets will be merged after changing column names to more consistent 
names (# 2 above) and formating (# 3 above). The data will be viewed 
chronologically after merging to look for inconsistencies in the data. 





































































































