---
title: "U.S. States' Disease Death Trends (2014-2021)"
author: "Patrick Cook"
date: "7/21/2021"
output: 
  html_document:
    df_print: paged
    keep_md: True
---

```{r, global.options, include = FALSE}
knitr::opts_chunk$set(
#    fig.width   = 12,          # the width for plots created by code chunk
#    fig.height  = 12,          # the height for plots created by code chunk
    dpi         = 200,         # set dpi for hd screen
    fig.align   = 'center',    # align graphics in the final doc
    out.width   = "90%",       # Set output (html) to % of window
    out.height  = "90%",       # Set output (html) to % of window
    # fig.path    = 'images/', # file path to graphics storage directory
    echo        = FALSE,       # do not display code in the code chunk
    message     = FALSE,       # do not display any messages generated by code
    strip.white = TRUE,        # remove white spaces at beg or end of code chunk
    warning     = FALSE,       # do not display any warning messages
    results     = TRUE)       # show text output
```


```{r , include=FALSE}
# Install packages
pacman::p_load(pacman, data.table, dplyr, tidyr, knitr, ggplot2, psych, cowplot, 
               corrplot, stringr, Hmisc, RColorBrewer)

# -----------------------------------------------------------------------------
# -------------- Set color palettes and sorting/helper variables --------------
# -----------------------------------------------------------------------------

# Diverging Palette: Modified from Brewer"Dark2", "Set1" and "Paired" palettes.
mycolors <- c("#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#00aaff", "#E6AB02", 
              "#A6761D", "#666666", "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", 
              "#FF7F00", "#FFFF33", "#A65628", "#F781BF", "#999999", "#66A61E", 
              "#B2DF8A", "#33A02C", "#FB9A99", "#E31A1C", "#FDBF6F", "#FF7F00", 
              "#CAB2D6", "#6A3D9A", "#FFFF99", "#B15928")

# create list of system greys (every other shade)
mygreycolors <- sprintf('grey%s', seq(0, 100, 2))

# Order string years [add current years as included]
ordered_years <- c("2014", "2015", "2016",
                   "2017", "2018", "2019", "2020", "2021")

# Some years are given 53 weeks in the year (usually leap years)
ordered_weeks <- seq(1:53)

# Order of states and special regions by population, greatest to least.
# Used for viewing states with similar population sizes and graphic ordering
state_by_pop <- c("California", "Texas", "Florida", "New York", "Pennsylvania", 
                  "Illinois", "Ohio", "Georgia", "North Carolina", "Michigan", 
                  "New Jersey", "Virginia", "New York City", "Washington", 
                  "Arizona", "Massachusetts", "Tennessee", "Indiana", 
                  "Missouri", "Maryland", "Wisconsin", "Colorado", "Minnesota", 
                  "South Carolina", "Alabama", "Louisiana", "Kentucky", 
                  "Oregon", "Oklahoma", "Connecticut", "Utah", "Puerto Rico", 
                  "Iowa", "Nevada", "Arkansas", "Mississippi", "Kansas", 
                  "New Mexico", "Nebraska", "West Virginia", "Idaho", "Hawaii", 
                  "New Hampshire", "Maine", "Montana", "Rhode Island", 
                  "Delaware", "South Dakota", "North Dakota", "Alaska", 
                  "District of Columbia", "Vermont", "Wyoming"
                  )

# End helper Variables

```

```{r , include=FALSE}
# -----------------------------------------------------------------------------
# --------------------------------- Functions ---------------------------------
# -----------------------------------------------------------------------------



# End Functions

```

#### **Note**: This file contains initial data cleaning and wrangling steps only. 
To view the analysis see:
1. [us_death_trends_analysis](us_death_trends_analysis.html)
2. [state_death_trends_analysis](state_death_trends_analysis.html)

### Introduction and Objectives

This report explores the Center for Disease Control's (CDC) Weekly Morbidity and 
Mortality data from 2014 through the 1st quarter of 2021. 
The data includes weekly death counts for specific causes in the United States. 
The data also includes weekly death counts for individual states and specific 
areas such as New York City, Puerto Rico and Washington, D.C. The data is 
voluntary and not guaranteed to be reported in a timely or regular basis. 
Therefore, the most recent data (quarter) will be incomplete and unreliable for 
analysis and insights. More information on the data set can be found at CDC's 
website using the following links: 

- [2014-2015 MMWR Data Set](https://data.cdc.gov/NCHS/Weekly-Counts-of-Deaths-by-State-and-Select-Causes/3yf8-kanr){target="_blank"}
- [2020-2021 MMWR Data Set](https://data.cdc.gov/NCHS/Weekly-Provisional-Counts-of-Deaths-by-State-and-S/muzy-jte6){target="_blank"}

Note: The most current data can be downloaded using the links above. To use the 
data sets without any code modifications, the user will need to:

1. Place the data two sets in a subdirectory named "data", and
2. Rename the data sets as "weekly_2014_2019" and "weekly_2020_2021."


```{r , include=FALSE}

# import data sets (included in data folder)

tmp_mmwr_1419_df = read.csv("data/weekly_2014_2019.csv",
                   header = TRUE,
                   stringsAsFactors = FALSE)

tmp_mmwr_2021_df = read.csv("data/weekly_2020_2021.csv",
                   header = TRUE,
                   stringsAsFactors = FALSE)

```

<br />  

#### Objective

The objective of the project is to analyze national and state level weekly cause
of death data available from the Centers for Disease Control (CDC) to identify 
specific diseases with the highest **growth** in deaths so that the information
may guide long-term and short-term planning. This planning will allow healthcare
systems to redirect resources, reducing financial, material, facility and labor 
cost wastage as well as identifying and addressing the cause for the increased 
deaths before the disease becomes a leading contributor of deaths.

### Initial Data Structure Exploration

```{r }
str(tmp_mmwr_1419_df)

```



```{r }
str(tmp_mmwr_2021_df)

```

<br />  

#### 2014-2019 data set head  

```{r }
head(tmp_mmwr_1419_df, 3)
```

<br />  

#### 2020-2021 data set head  

```{r }
head(tmp_mmwr_2021_df, 3)
```


<br />  

#### Check flags types 

```{r }
# Verify flag column entries are only 'Suppressed (counts1-9)' privacy flags.
unique(unlist(select(tmp_mmwr_1419_df, contains('flag'))))
unique(unlist(select(tmp_mmwr_2021_df, contains('flag'))))
``` 

<br />  

#### Feature name differences  

Columns in 2014-2019 data **missing from 2020-2021 data**  

```{r }
# Differences in feature names of two data sets

tmp_col_list_1419 <- colnames(tmp_mmwr_1419_df)
tmp_col_list_2021 <- colnames(tmp_mmwr_2021_df)

print('---- Columns in 2014-2019 data missing from 2020-2021 data')
tmp_col_list_1419[!(tmp_col_list_1419 %in% tmp_col_list_2021)]

```

<br />  

Columns in 2020-2021 data **missing from 2014-2019 data**  

```{r }
tmp_col_list_2021[!(tmp_col_list_2021 %in% tmp_col_list_1419)]

```

<br />  

#### Data Structure Observations and changes needed  

1.  The 2014-2019 data set has 16,902 records with 30 features. The 2020-2021 
data contains 4,428 records with 35 features. The difference in features is due 
to the addition of two Covid-19 categories in the 2020-2021 data set.

2. Both data sets have flag columns that identify weekly death counts that are 
from 1 to 9 deaths that have been redacted due to privacy issues. The records 
with missing values are currently all due to the privacy flag. These flags 
columns are not needed for the analysis of growth rates and will be dropped. The
missing values will not be replaced due to the influence a zero or interpolated 
value would have on statistical calculations and patterns seen in plots. The 
missing values do still have value in that they indicate a positive weekly death 
count below 10. 

2. Both data sets contain coded feature names based on International Classification 
of Diseases (ICD-10) that will be renamed to shorter names based on affected 
organ or two word disease name.

3. The dates in both data sets are listed as characters and are not in the 
global standard format. Data type will be changed and put into ISO 8601 format.

4. The jurisdictions will be renamed to "Location" and changed to a factor.

5. Influenza.and.pneumonia has and additional ICD-10 code (J09) in the 2020-2021
data. ICD-10 code J09 was added for classifying specific strains of Influenza.

6. Data.As.Of exists in the 2020-2021 data set and is used to track updates. The
update information is a last update and not an indication of record completeness 
therefore this feature will be dropped. 

7. All..Cause vs All.Cause is a typing difference. This includes non-disease 
deaths and will not be used in the analysis. Therefore, the column will be 
removed.

8. Other features such as year and week may be changed to ordered factors as 
needed.

Both data sets will be merged after changing column names to more consistent 
names (# 2 above) and formatting (# 3 above). The data will be viewed 
chronologically after merging to look for inconsistencies in the data. 


```{r }
# Cleaning and Wrangling processing

# Dropping flag columns

tmp_mmwr_1419_df <- select(tmp_mmwr_1419_df,
                           -contains(c('flag', 'All.')))

tmp_mmwr_2021_df <- select(tmp_mmwr_2021_df, 
                           -contains(c('flag', 'Data.As.Of', 'All.')))

# Rename feature list to more readable and analysis-friendly names

col_list <- c('Location', 'Year', 'Week', 'Week_End_Date', 'Natural', 
              'Septicemia', 'Cancer', 'Diabetes', 'Alzheimer', 
              'Influenza_Pneumonia', 'Lower_Respiratory', 'Other_Respiratory', 
              'Kidney', 'Abnormal_Finding', 'Heart', 'Brain', 'Covid_19_Multi', 
              'Covid_19')

# Add Covid-19 features filled with 0 and update feature names for 2014-19 data
tmp_mmwr_1419_df[col_list[17:18]] <- 0

# Convert Covid-19 features to integer from number
tmp_mmwr_1419_df$Covid_19_Multi <- as.integer(tmp_mmwr_1419_df$Covid_19_Multi)
tmp_mmwr_1419_df$Covid_19 <- as.integer(tmp_mmwr_1419_df$Covid_19)

# Update feature names to common names prior to merging
colnames(tmp_mmwr_1419_df) <- col_list
colnames(tmp_mmwr_2021_df) <- col_list

# Change data types and formatting for dates
tmp_mmwr_1419_df$Week_End_Date <- as.Date(tmp_mmwr_1419_df$Week_End_Date,
                                          format='%m/%d/%Y')

tmp_mmwr_2021_df$Week_End_Date <- as.Date(tmp_mmwr_2021_df$Week_End_Date,
                                          format='%Y-%m-%d')

```


```{r }
# Merge data sets, clean global environment and confirm changes

if (exists("tmp_mmwr_1419_df") && exists("tmp_mmwr_2021_df")) {
  
  # Merge data sets into data.table
  mmwr_1421_df <- rbindlist(list(tmp_mmwr_1419_df, tmp_mmwr_2021_df))
  
  # Change Location to factors
  mmwr_1421_df$Location <- as.factor(mmwr_1421_df$Location)

  # Remove temporary data frames, environmental variables
  rm(list = ls(pattern = "^tmp_"))
}

```

```{r }
# Create ordered causes of death vector from greatest to least.

# Initial order of disease columns
tmp_char_vector <- c('Natural', 'Septicemia', 'Cancer', 'Diabetes', 
                           'Alzheimer', 'Influenza_Pneumonia', 
                           'Lower_Respiratory', 'Other_Respiratory', 'Kidney', 
                           'Abnormal_Finding', 'Heart', 'Brain', 
                           'Covid_19_Multi', 'Covid_19')

# Get means of National disease deaths counts
# Order diseases from greatest deaths to least and add to col_list
deaths_ordered <- tmp_char_vector[rev(
                                  order(
                                  colMeans(
                                   mmwr_1421_df[Location == 'United States',
                                                Natural:Covid_19],
                                   na.rm=TRUE)))]

col_list <- c('Location', 'Year', 'Week', 'Week_End_Date', deaths_ordered)

rm(list = ls(pattern = "^tmp"))

# reorder table by col_list (most significant deaths to least)
mmwr_1421_df <- mmwr_1421_df[ , col_list, with = FALSE]

# Export cleaned data set
write.csv(mmwr_1421_df,
          "data\\processed_data\\mmwr_1421_df.csv",
          row.names = FALSE)

``` 

### Data set after initial Cleaning and Wrangling  

```{r}
# Verify Changes
head(mmwr_1421_df)

```

<br />  

#### **Note on Analysis**:
The analysis of the national data will be performed separately from the state 
data. To view the analysis see:  

1. [us_death_trends_analysis](us_death_trends_analysis.html)
2. [state_death_trends_analysis](state_death_trends_analysis.html)  

<br />  
